p-adic precision

### 2014-01-29 padicprec.tex
Proof of the correctness of Algorithm SOMOS

%\subsection{Differential tracking of precision}
%
%\todo{
%Explain here that we can track precision by computing differentials step 
%by step. Note also in this subsection that this method can be very time 
%consuming; it's why we often prefer to consider other models of precision.
%}

@@

\subsection{Polynomials}

\todo{
Euclidean division, factorization, \emph{etc.}
}

\subsection{Matrices}

\todo{
LU factorization, determinant, characteristic polynomial, kernels, images,
\emph{etc.}
}

@@

In this last paragraph, we illustrate the theory developed in this 
article by designing an algorithm for computing the $u_n$'s at precision 
$O(p^N)$. The algorithm is presented at the top of page 
\pageref{algo:somos}. We now prove that it is correct.
%
\begin{algorithm}[t]
\SetKwInOut{Assumption}{Assumption}
\KwIn{$a,b,c,d$ --- four initial terms of a SOMOS 4 sequence $(u_n)_{n \geq 0}$}
\KwIn{$n, N$ --- two integers}
\Assumption{$a$, $b$, $c$ and $d$ lie in $\Z_p^\times$ and are known at 
precision $O(p^N)$}
\Assumption{None of $u_i$ $(0 \leq i \leq n$) is divisible by $p^N$}
\KwOut{$u_n$ at precision $O(p^N)$}
\BlankLine
$\prec$ $\leftarrow$ $N$\;
\For {$i$ from $1$ to $n-3$}{
  $\prec$ $\leftarrow$ $\prec + v_p(bd + c^2) - v_p(a)$\;
  {\bf lift} (arbitrary) 
    $b$, $c$ and $d$ to precision $O(p^{\prec})$\;
  $\prec$ $\leftarrow$ $\prec - v_p(a)$\;
  $e$ $\leftarrow$ $\frac{bd+c^2} a$; \hspace{1cm}
   \tcp{$e$ is known at precision $O(p^\prec)$}
  $a, b, c, d$ $\leftarrow$ $b + O(p^{\prec}), c + O(p^{\prec}), d + O(p^{\prec}), e$\;
}
\Return{$d + O(p^N)$}\;
\caption{\sc SOMOS$(a, b, c, d, n, N)$}\label{algo:somos}
\end{algorithm}
%
We introduce the (partial) function $f : \Q_p^4 \to \Q_p^4$ defined by 
$f(a,b,c,d) = (b,c,d,\frac{bd+c^2}a)$, so that, for all $i$, we have 
$(u_i, u_{i+1}, u_{i+2}, u_{i+3}) = f_i(u_0, u_1, u_2, u_3)$ where $f_i
= f \circ \cdots \circ f$ ($i$ times). Apparently $f$ is strictly 
differentiable on $\Q_p^\star \times \Q_p^3$ and its differential in the 
canonical basis is given by the matrix:
$$D(a,b,c,d) = \begin{pmatrix}
0 & 1 & 0 & 0 \\
0 & 0 & 1 & 0 \\
0 & 0 & 0 & 1 \\
-\frac{bd+c^2}{a^2} & \frac d a & \frac {2c} a & \frac b a
\end{pmatrix}$$
whose determinant is $\frac{bd+c^2}{a^2}$. Thus, if the $(i+4)$-th term 
of the SOMOS sequence is defined, the mapping $f_i$ is differentiable
at $(u_0, u_1, u_2, u_3)$ and the matrix its differential at this point 
is:
$$D_i = D(u_{i-1}, u_i, u_{i+1}, u_{i+2}) \cdots
D(u_1, u_2, u_3, u_4) \cdot D(u_0, u_1, u_2, u_3).$$
%Its determinant is:
%$$\det D_i = \frac{u_{i+3}}{u_{i-1}} \cdots 
%\frac{u_5}{u_1} \cdot \frac{u_4}{u_0} =
%\frac{u_i \cdot u_{i+1} \cdot u_{i+2} \cdot u_{i+3}}
%{u_0 \cdot u_1 \cdot u_2 \cdot u_3}.$$
Moreover thanks to the Laurent phenomenon, we know that $D_i$ has
integral coefficients. Hence, letting $\varphi_i = f'_i(u_0, u_1, u_2, 
u_3)$, we get 
$$(\det D_i) \cdot p^N \Z_p^4 \subset \varphi_i(p^N \Z_p^4) 
\subset p^N \Z_p^4$$
for all integer $N$.

@@

%ACKNOWLEDGMENTS are optional
\section{Acknowledgments}

@@

\bibitem{Bostan}
{\sc Bostan, Alin, \& al.} 
\newblock From Newton sums to coefficients: complexity issues in characteristic p. Proceedings MEGA'05, 2005.

\bibitem{Caruso}
{\sc Caruso, X.}
\newblock Random matrices over a DVR and LU factorization, preprint (2012)

\bibitem{Gaudry}
{\sc Gaudry,P. \& al.} 
\newblock The 2-adic CM method for genus 2, Asiacrypt 2006, volume 4284 of Lecture Notes in Comput. Sci., pages 114–129. Springer, Berlin, 2006.

\bibitem{Kedlaya}
{\sc Kedlaya,Kiran,} 
\newblock Counting points on hyperelliptic curves using Monsky--Washnitzer cohomology, Journal of the Ramanujan Mathematical Society 16 (2001),

\bibitem{Lercier}
{\sc Lercier, Reynald \&  Sirvent, Thomas,} 
\newblock On elkies subgroups of $\ell$-torsion points in elliptic curves defined over a finite field. Journal de Théorie des Nombres de Bordeaux, 20(3):783-797, December 2008.

\bibitem{Schneider}
{\sc Schneider, Peter}
\newblock $p$-adic Lie Groups, Springer, 2011

\end{thebibliography}

>>>>

@@

\subsection{Differential of usual operations}

\subsubsection{Numbers}

\todo{sum, product, inverse, $p$-power map}

\subsubsection{Polynomials}

\todo{
Euclidean division, factorization, Grobner basis (?)
}

\subsubsection{Matrices}

\todo{
determinant, characteristic polynomial, LU factorization, QR 
factorization, kernels, images}

\subsubsection{Sub vector spaces}

\todo{intersection, sum}

@@

In this last paragraph, we illustrate the theory developed in this 
article by designing an algorithm for computing the $u_n$'s at precision 
$O(p^N)$. The algorithm is presented at the top of page 
\pageref{algo:SOMOS}. We now prove that it is correct.
%
\begin{algorithm}[t]
\SetKwInOut{Assumption}{Assumption}
\KwIn{$a,b,c,d$ --- four initial terms of a SOMOS 4 sequence $(u_n)_{n \geq 0}$}
\KwIn{$n, N$ --- two integers}
\Assumption{$a$, $b$, $c$ and $d$ lie in $\Z_p^\times$ and are known at 
precision $O(p^N)$}
\Assumption{None of $u_i$ $(0 \leq i \leq n$) is divisible by $p^N$}
\KwOut{$u_n$ at precision $O(p^N)$}
\BlankLine
$\prec$ $\leftarrow$ $N$\;
\For {$i$ from $1$ to $n-3$}{
  $\prec$ $\leftarrow$ $\prec + v_p(bd + c^2) - v_p(a)$\;
  {\bf lift} (arbitrary) 
    $b$, $c$ and $d$ to precision $O(p^{\prec})$\label{line:lift}\;
  $\prec$ $\leftarrow$ $\prec - v_p(a)$\;
  $e$ $\leftarrow$ $\frac{bd+c^2} a$; \hspace{1cm}
   \tcp{$e$ is known at precision $O(p^\prec)$}
  $a, b, c, d$ $\leftarrow$ $b + O(p^{\prec}), c + O(p^{\prec}), d + O(p^{\prec}), e$\;
}
\Return{$d + O(p^N)$}\;
\caption{\sc SOMOS$(a, b, c, d, n, N)$}\label{algo:SOMOS}
\end{algorithm}
%
We introduce the (partial) function $f : \Q_p^4 \to \Q_p^4$ defined by 
$f(a,b,c,d) = (b,c,d,\frac{bd+c^2}a)$, so that, for all $i$, we have 
$(u_i, u_{i+1}, u_{i+2}, u_{i+3}) = f_i(u_0, u_1, u_2, u_3)$ where $f_i
= f \circ \cdots \circ f$ ($i$ times). Apparently $f$ is strictly 
differentiable on $\Q_p^\star \times \Q_p^3$ and its differential in the 
canonical basis is given by the matrix:
$$D(a,b,c,d) = \begin{pmatrix}
0 & 1 & 0 & 0 \\
0 & 0 & 1 & 0 \\
0 & 0 & 0 & 1 \\
-\frac{bd+c^2}{a^2} & \frac d a & \frac {2c} a & \frac b a
\end{pmatrix}$$
whose determinant is $\frac{bd+c^2}{a^2}$. Thus, if the $(i+4)$-th term 
of the SOMOS sequence is defined, the mapping $f_i$ is differentiable
at $(u_0, u_1, u_2, u_3)$ and its differential $\varphi_i = 
f'_i(u_0, u_1, u_2, u_3)$ is given by the matrix:
$$D_i = D(u_{i-1}, u_i, u_{i+1}, u_{i+2}) \cdots
D(u_1, u_2, u_3, u_4) \cdot D(u_0, u_1, u_2, u_3).$$
Moreover thanks to the Laurent phenomenon, we know that $D_i$ has
integral coefficients, \emph{i.e.} $\varphi_i$ stabilizes the lattice
$\Z_p^4$.
We are now going to prove by induction on $i$ that, at the end of the 
$i$-th iteration of the loop, we have $\prec = N + v_p(\det D_i)$ and
\begin{equation}
\label{eq:congrSOMOS}
(a, b, c, d) \equiv (u_i, u_{i+1}, u_{i+2}, u_{i+3}) \pmod
{p^N \varphi_i(\Z_p^4)}.
\end{equation}
The first point is easy. Indeed, from $D_i = D(u_{i-1}, u_i, u_{i+1}, 
u_{i+2}) D_{i-1}$, we deduce $\det D_i = \det D_{i-1} \cdot 
\frac{u_i}{u_{i-3}}$ and the assertion follows by taking determinants
and using the induction hypothesis. 
Let us now establish Eq.~\eqref{eq:congrSOMOS}. To avoid confusion, 
let us agree to denote $a'$, $b'$, $c'$, $d'$ and $\prec'$ the values of 
$a$, $b$, $c$, $d$ and $\prec$ respectively at the \emph{beginning} of 
the $i$-th iteration of the loop. By induction hypothesis (or by 
initialization if $i = 1$), we have:
\begin{equation}
\label{eq:congrSOMOS2}
(a', b', c', d') \equiv (u_{i-1}, u_i, u_{i+1}, u_{i+2}) \pmod
{p^N \varphi_{i-1}(\Z_p^4)}.
\end{equation}
Moreover, we know that the determinant of $\varphi_{i-1}$ has valuation
$\prec'$. Hence \eqref{eq:congrSOMOS2} remains true if $a'$, $b'$, $c'$
and $d'$ are replaced by other values which are congruent to them modulo
$p^{\prec'}$. In particular it holds if $a'$, $b'$, $c'$ and $d'$ denotes
the values of $a$, $b$, $c$ and $d$ after the execution of line
\ref{line:lift}. Now applying Lemma \ref{lem:main} and Proposition
\ref{prop:} to $\varphi_{i-1}$ and $\varphi_i$ (at the point $(u_0,
u_1, u_2, u_3)$), we get:
$$f\big((u_{i-1}, u_i, u_{i+1}, u_{i+2}) + p^N \varphi_{i-1}(\Z_p^4)\big)
= (u_i, u_{i+1}, u_{i+2}, u_{i+3}) + p^N \varphi_i(\Z_p^4).$$
By the discussion above, this implies in particular that $f(a',b',c',d')$
belongs to $(u_i, u_{i+1}, u_{i+2}, u_{i+3}) + p^N \varphi_i(\Z_p^4)$. We
conclude by remarking first that $(a,b,c,d) \equiv f(a',b',c',d') \pmod 
{p^\prec \Z_p^4}$ by construction and second that $p^\prec \Z_p^4 \subset
p^N \varphi_i(\Z_p^4)$.

Finally Eq.~\eqref{eq:congrSOMOS} applied with $i = n-3$ together with
the fact that $\varphi_i$ stabilizes $\Z_p^4$ imply that, when we exit
the loop, the value of $d$ is congruent to $u_n$ modulo $p^N$. Hence,
our algorithm returns the correct value.

\medskip

Let us conclude this subsection by remarking that Algorithm 
\ref{algo:SOMOS} performs computations at precision at most $O(p^{N+v})$ 
where $v$ is the sum of the valuations of five consequence terms among 
the first $n$ terms of the SOMOS sequence we are considering. 
Experiments show that the value of $v$ varies like $c \cdot \log n$ 
where $c$ is some constant. Assuming that we are using a FFT-like 
algorithm to compute products of integers, the complexity of Algorithm 
\ref{algo:SOMOS} is then expected to be $\tilde O(N n)$ where the 
notation $\tilde O$ means that we hide logarithmic factors.

We can compare this with the complexity of the more naive algorithm 
consisting of lifting the initial terms $u_0, u_1, u_2, u_3$ to enough 
precision and then do the computation using a naive step-by-step 
tracking of precision. In this setting, the required original precision 
is $O(p^{N+2v'})$ where $v'$ is the sum of the valuation of the $u_i$'s 
for $i$ varying between $0$ and $n$. Experiments show that $v'$ is about 
$c' \cdot n \log n$ (where $c'$ is a constant), which leads to a 
complexity in $\tilde O (N n + n^2)$. Our approach is then interesting
when $n$ is large compared to $N$: under this hypothesis, it saves
roughly a factor $n$.

@@

@@

\bibitem{Bostan}
  A.~Bostan, L.~Gonz\'alez-Vega, H.~Perdry, \'E.~Schost,
  \emph{From Newton sums to coefficients: complexity issues in characteristic $p$},
  Proceedings MEGA'05 (2005).

\bibitem{Caruso}
  X.~Caruso,
  \emph{Random matrices over a DVR and LU factorization},
  preprint (2012)

\bibitem{Gaudry}
  P.~Gaudry, T.~Houtmann, A.~Weng, C.~Ritzenthaler, D.~Kohel,
  \emph{The $2$-adic CM method for genus $2$}, 
  Asiacrypt 2006, vol. 4284 of Lecture Notes in Comput. Sci. (2006), 114--129

\bibitem{Kedlaya}
  K.~Kedlaya,
  \emph{Counting points on hyperelliptic curves using Monsky--Washnitzer cohomology}, 
  J. Ramanujan Math. Soc. {\bf 16} (2001)

\bibitem{Lercier}
  R.~Lercier, T.~Sirvent,
  \emph{On Elkies subgroups of $\ell$-torsion points in elliptic curves defined over a finite field},
  J. Th\'eorie des Nombres de Bordeaux {\bf 20} (2008), 783--797

\bibitem{Schneider}
  P.~Schneider,
  \emph{$p$-adic Lie Groups}, Springer (2011)
\end{thebibliography}
